Peptide retention time prediction

### 2015-01-11 retention.tex
Small textual changes in the Discussion

The accuracy of the machine learning-based methods described in
section \ref{sec:ML} is still much lower than the methods based on
look ups of previous registered retention time. However, the large
availability of datasets leads one to think that there still is a
possibility to increase the accuracy of such predictions by enhanced
machinelearning technology.  We can see that there are a number of
interesting techniques that not yet have been used on the problem of
predicting peptides retention time. Here we will bring up one such
techniques, that we feel are particularly good matches, that we have
not yet seen implemented for the problem. Pairwise Machine-learned
ranking (MLR) \cite{liu2009learning} models can be trained to
determine which of two examples that should be better ranked thatn the
other. MLR is traditionally used for construction of ranking models
for information retrieval systems. However, in the case of retention
time prediction we can use such a predictor to predict the elusion
order for sets of peptides, which can be useful in its own
\cite{bailey2012instant}, but can also be converted into absolute
retention times, given that the distribution of their retention times
are known.

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that ideathe performance of predictors often are enhanced
if one can structure the prediction into multiple layers. The drawback
of such models are that they require large training sets. Given the
very large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas \cite{desiere2006peptideatlas} and
PRIDE \cite{martens2005pride}, such methods could probably be
implemented for retention time prediction as well.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.

>>>>

The accuracy of the machine learning-based methods described in
section \ref{sec:ML} is still much lower than the methods based on
look-ups of previously registered retention times. However, the large
availability of datasets suggests that one possibility to increase the
accuracy of retention time predictions could rely on the use of
enhanced machine learning methods. Currently, there are a number of
such techniques that have not yet been applied to the problem of
predicting peptides' retention time. Here we will bring up one such
technique called pairwise Machine-Learned Ranking (MLR)
\cite{liu2009learning}, which we feel is particularly suitable for
this problem. MLR models can be trained to determine which of two data
points should be better ranked.

MLR has traditionally been used for construction of ranking models for
information retrieval systems. However, in the case of retention time
prediction, we can use such a predictor to estimate the elution order
for sets of peptides. While such information can be useful in its own
\cite{bailey2012instant}, it can also be converted into absolute
retention times, given that the distribution of the retention times is
known.

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models are that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM), are dependent on the predictability of a
peptide's behavior. In such experiments the retention times and the
suitable transitions are needed as input to the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiments
would lead to a more direct experimental design and hopefully to more
comprehensie coverages. Thus {\em de novo} retention time predictions
have an important role to fill.
